biological data preperation

### 2018-09-04 src/biological_data_prep.Rmd
to fix missing inst_code

#Overview
Preperation of the capelin (*Mallotus villosus*) data obtained from [OBIS](www.iobis.org). This database shows the global occurance of capelin. 
Built with 'r getRversion()'.

#Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
Th
```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

#Raw Data Clean
Load the raw data, then check use head() to check it looks ok, then names() to see all the column names
```{r load raw data}
rawdata <- read.csv("../data/bio/capelin_obis_raw.csv")
head(rawdata) 
names(rawdata)
```

@@

The dataset is global and needs to be trimmed to contain just the area of interest. 
Plot all the occurance points
```{r map all occurrence points}
library(rworldmap)
map <- getMap(resolution = "low") #creates an object called map at medium resoultion
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #the x and y lim are the long-lat bounds of the map
points(tidydata$decimalLongitude, tidydata$decimalLatitude) #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
```
now extract just the Atlantic Canada data.
The lon-lat study area (which captures the whole Atlantic Canadian EEZ up to the north of Resolution Island in the Hudson Strait/ Davis Strait) is -71.0,-43.0,62.0,38.0 
```{r map only atlantic canadian occurrences}
tidydata <- tidydata[tidydata$decimalLongitude > -71, ]
tidydata <- tidydata[tidydata$decimalLongitude < -43, ]
tidydata <- tidydata[tidydata$decimalLatitude > 38, ]
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #re-map - this is just atl. canada (after checked subsetting is good)
points(tidydata$decimalLongitude, tidydata$decimalLatitude)
```
ok looks good...save what you have so far to a .csv
```{r}
write.csv(tidydata, file = "../data//bio/capelin_obis_tidy.csv", row.names = FALSE)
```

>>>>

# Overview
Preperation of the capelin (*Mallotus villosus*) data obtained from [OBIS](www.iobis.org). This database shows the global occurance of capelin. 
Built with 'r getRversion()'.

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
Th
```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

# Raw Data Clean
Load the raw data, then check use head() to check it looks ok, then names() to see all the column names
```{r load raw data}
rawdata <- read.csv("../data/bio/capelin_obis_raw.csv")
head(rawdata) 
names(rawdata)
```

@@

The dataset is global and needs to be trimmed to contain just the area of interest. 
Plot all the occurance points
```{r map all occurrence points}
library(rworldmap)
map <- getMap(resolution = "low") #creates an object called map at medium resoultion
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #the x and y lim are the long-lat bounds of the map
points(tidydata$decimalLongitude, tidydata$decimalLatitude) #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
```
now extract just the Atlantic Canada data.
The lon-lat study area (which captures the whole Atlantic Canadian EEZ up to the north of Resolution Island in the Hudson Strait/ Davis Strait) is -71.0,-43.0,62.0,38.0 
```{r map only atlantic canadian occurrences}
tidydata <- tidydata[tidydata$decimalLongitude > -71, ]
tidydata <- tidydata[tidydata$decimalLongitude < -43, ]
tidydata <- tidydata[tidydata$decimalLatitude > 38, ]
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #re-map - this is just atl. canada (after checked subsetting is good)
points(tidydata$decimalLongitude, tidydata$decimalLatitude)
```
ok looks good...save what you have so far to a .csv
```{r save tidydata to csv}
write.csv(tidydata, file = "../data//bio/capelin_obis_tidy.csv", row.names = FALSE)
```

# Column removal
So there are many columns that you don't need.. you want to keep
id
decimalLatitude
decimalLongitude
datecollected
institutioncode
individualcount
resname
originalscientificname

```{r select colums needed}
tidydata <- subset(tidydata, select = c(id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, resname, originalscientificname))
head(tidydata)
```

# remove data without month/year info

count NAs in the dataset. Mainly you are interested in the datecollected column (ignore individual count - you know there is a lot missing)
```{r}
missingdata <- apply(tidydata, 2, function (x) sum(is.na(x)))
missingdata
```
hmm interesting - you have 771 columns from year month and day with missing data, but datacollected has 0. There may be a formatting issue.
To take a closer look, subset all rows where year == na

```{r}
missingdatarows <- tidydata[is.na(tidydata$year), ]
head(missingdatarows)
```
so the NAs in the year column relate to no data (without a na value attached). Remove all rows where year == na
```{r}
tidydata <- tidydata[!is.na(tidydata$year), ]
missingdata2 <- apply(tidydata, 2, function (x) sum(is.na(x)))
missingdata2
```

# Month year column creation
make columns for the years and month based on the datecollected column
```{r new columns}
tidydata$year <- format(as.Date(tidydata$datecollected), "%Y") #created a new column called year
tidydata$month <- format(as.Date(tidydata$datecollected), "%m") #created a new column called month
tidydata$day <- format(as.Date(tidydata$datecollected), "%d") #created a new column called day 
head(tidydata)
```

the year, month, day columns are characters Change to numeric
```{r new columns 2}
tidydata$year <- as.numeric(tidydata$year)
tidydata$month <- as.numeric(tidydata$month)
tidydata$day <- as.numeric(tidydata$day)
head(tidydata)
```

# Institutions
Lets take a look at the institutioncode to make sure the names make sense...
```{r}
tidydata$institutioncode <- as.character(tidydata$institutioncode)
whosampled <- unique(tidydata$institutioncode) #makes a variable of institutions
whosampledordered <- sort(whosampled) #just puts the list in 
whosampledordered
```
Ok so a number of these names actually represent the same insitiution (e.g. [DFO Regions](http://www.dfo-mpo.gc.ca/regions/index-eng.htm). Some details [here](ftp://ftp.aoml.noaa.gov/od/pub/library/methods-branton.pdf) and the first is no institution (you might be able to figure it out though). 
The merges needed are as follows
"Bedford Institute of Oceanography (BIO)" & "BIO" & "DFO-SF" == DFOMTMS 
"DFO-IML" & "IML-MLI" & "IML_MLI" & "DFO-NG" == DFOQC 
"DFO-NFLD" & "North Atlantic Fisheries Centre" == "DFONL" 
"NMFS-NEFSC" & "NOAA, NMFS, Northeast Fisheries Science Center" == NEFSC
"The Atlantic reference Centre" & "The Atlantic Reference Centre (ARC)" == ARC 
"DFO-SG" & "DFO Gulf Fisheries Centre" == DFOGULF

and some you can just make shorter
"Northwest Atlantic Fisheries Organization (NAFO)" == NAFO
"DFO-ISDM" == DFOISDM
"Maine DMR" == MaineDMR
```{r}
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "Bedford Institute of Oceanography (BIO)", "DFOMTMS")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "BIO", "DFOMTMS")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-SF", "DFOMTMS")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-IML", "DFOQC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "IML-MLI", "DFOQC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-NG", "DFOQC")  
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "IML_MLI", "DFOQC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-NFLD", "DFONL")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "North Atlantic Fisheries Centre", "DFONL")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "NMFS-NEFSC", "NEFSC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "NOAA, NMFS, Northeast Fisheries Science Center", "NEFSC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "The Atlantic Reference Centre (ARC)", "ARC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "The Atlantic reference Centre", "ARC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-SG", "DFOGulf")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO Gulf Fisheries Centre", "DFOGulf")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "Northwest Atlantic Fisheries Organization (NAFO)", "NAFO")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-Central and Arctic, Freshwater Institute (FWI)", "DFOCENARC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-ISDM", "DFOISDM")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "Maine DMR", "MaineDMR")

#recheck
tidydata$institutioncode <- as.character(tidydata$institutioncode)
whosampled <- unique(tidydata$institutioncode) #makes a variable of institutions
whosampledordered <- sort(whosampled) #just puts the list in 
whosampledordered
```

Now look at the missing institution. Use resname to see if it helps...
```{r}
missinginst <- subset(tidydata, institutioncode == "")
missinginst
```
ok now check for unique resname...
```{r}
missinginst$resname <- as.character(missinginst$resname)
missinginstunique <- unique(tidydata$resname) #makes a variable of institutions
missinginstuniquedordered <- sort(missinginstunique) #just puts the list in 
missinginstuniquedordered <- as.character(missinginstuniquedordered)
missinginstuniquedordered
```
Ugh ok there are quite a few here....

Ok soo looking at the OBIS metadata [e.g.](http://www.iobis.org/explore/#/dataset/2327), this is what they should be

Arctic Species Trend Index (ASTI) : Marine == CAFF
Atlantic Reference Centre Museum of Canadian Atlantic Organisms - Invertebrates and Fishes Data == ARC
Atlantic Zone Monitoring Program Maritimes Region (AZMP) plankton datasets. In: Fisheries and Oceans Canada - BioChem archive == DFOMTMS
BioChem zooplankton samples from the Gully, 2006-2007 == DFOMTMS
BioChem:  Zooplankton of St. Margarets Bay 1968 to 1971. == DFOMTMS
BioChem: DFO Atlantic Zone Monitoring Program plankton datasets - Newfoundland and Labrador Region (OBIS Canada) ==  DFONL
BioChem: Fish, Eggs and Larvae in the Gulf of St. Lawrence == DFOQC
BioChem: Sameoto zooplankton collection == DFOMTMS 
DFO Central and Arctic Multi-species Stock Assessment Surveys == DFOCENARC
DFO Gulf Region Groundfish Research Vessel Surveys == DFOGulf
DFO Maritimes Research Vessel Trawl Surveys Fish Observations == DFOMTMS
DFO Newfoundland and Labrador Region Ecosystem Trawl Surveys == DFOQC
DFO Quebec Region MLI museum == DFOQC
DFO Quebec Region Multispecies bottom trawl surveys == DFOQC
ECNASAP - East Coast North America Strategic Assessment == DFOMTMS
Electronic Atlas of Ichthyoplankton on the Scotian Shelf of North America == DFOMTMS
Fish specimens == ROM
ICES contaminants and biological effects == ICES
Ichthyology Collection - Royal Ontario Museum == ROM
IMR Macroplankton surveys == IMR
Maine Department of Marine Resources Inshore Trawl Survey 2000–2009 == MaineDMR
NAFO/ICNAF - Environmental Surveys - NORWESTLANT 1-3, 1963: Fish eggs and larvae == NAFO
NMNH Vertebrate Zoology Fishes Collections == NMNH
Northeast Fisheries Science Center Bottom Trawl Survey Data == NEFSC
Northern Gulf of St. Lawrence Fishes == DFOMTMS

ok so now you want to fill in the missing institutioncode with the correct data...

maybe just populate the missing values with the same text as resname
...
if institutioncode == ""
institutioncode == "resname"
```{r}
for i in (1:nrow)
```
